Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-06-02 write/manuscript.tex
added governing equations; definition of Reynolds number

\usepackage[style=authoryear-comp,firstinits=true,citestyle=authoryear,natbib=true,backend=bibtex]{biblatex}
\renewcommand\nameyeardelim{, }
\DeclareNameAlias{sortname}{family-given}
\renewcommand*{\multinamedelim}{\addcomma\space}
\renewcommand*{\finalnamedelim}{\addcomma\space}
\setlength{\bibitemsep}{\baselineskip}
\renewbibmacro{in:}{}

@@

\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}
\newcommand{\SFC}{\mathrm{sfc}}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
physical effects on large and small scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Locally, the directional shear of the wind in the upper part of the surface layer may cause a systmatic yaw
for tall wind power generation devices where blades reach into the Ekman layer, i.e. that part of the boundary
layer where the wind starts to turn. 
%
\par
%
Wind veer in the planetary boundary layer is commonly characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Therefore, knowing these parameters is a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \cite{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \cite{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \cite{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\cite{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \cite[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

\cite{gryning:BM2007} present an extension of the wind-speed profile beyond the surface layer

@@

\cite{jiang:JAS2018} recognized that the outer part of the Ekman boundary layer receives less attention in comparison

@@

While recent advances have been made in analysis of this simplified set-up \cite{jiang:JAS2018} (MORE?),
there is not enough understanding for a quantitative generalization of the results to arbitrary external forcing.

@@

\todo{Numerical approach, DNS, Scale-separation, Re-similarity
  \citep{dimotakis:JFM2000, dasilva:ARF2014, pope:2000, moin:ARF1998, mellado:JAM2018}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Formulation}
%
We consider here incompressible, turbulent Ekman flow, that is, the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
The f-plane approximation is applied such that rotation only acts on horizontal velocity components;
we thus neglect rotational effects on the horizontal components of velocity and dynamical effects
due to latitudinal variation of the rate of rotation.
%
\par
%
The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment (plus/ minus few degrees centigrade) of the direction $Ox$ with the surface shear stress is achieved.
The coordinate $Oy$ points away from the wall,  and $Oz$ points in the span-wise direction normal to $Oxy$. 
%
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity.

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}

Formulations for the outer layer that take into account the rotation, and thus deviation from the channel-flow analogy
need to be matched to the framework of surface similarity, and it
occurs that a smooth transition from the inner to the outer layer is not easily achieved.
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we believe that it is the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.

>>>>

\usepackage[maxcitenames=2,style=authoryear-comp,firstinits=true,citestyle=authoryear,natbib=true,backend=bibtex]{biblatex}
\DeclareNameAlias{sortname}{family-given}
\renewcommand*{\multinamedelim}{\addcomma\space}
\renewcommand*{\finalnamedelim}{~and\space}
\setlength{\bibitemsep}{\baselineskip}
\renewbibmacro{in:}{}

@@

\newcommand{\todo}[1]{\textcolor{red}{$[$#1$]$}}

% TO BE USED IN MATH-ENVIRONMENT 
\newcommand{\SFC}{\mathrm{sfc}}
\newcommand{\p}{\partial}
\newcommand{\gzm}{{\xi}}%{\gamma z_\text{match}}}
\newcommand{\RO}{\mathrm{Ro}}  
\newcommand{\RE}{\mathrm{Re}}
\newcommand{\LR}{\Lambda_\RO}

@@

Wind veer away from its geostrophic direction is
(i) due to direct frictional effects in the very vicinity of the surface and
(ii) due to turbulence which exerts indirect frictional effects; these effects
cause a slow-down of the mean wind reducing the Coriolis force thus turning the wind
in favor of the pressure gradient force. 
%%
Not only does the veering set the frame of reference for surface layer theory, it also has
physical effects on large and small scales.
%
From a large-scale perspective, the veering of wind across the planetary boundary layer determines
the amount of cross-isobaric mass-flux, commonly referred to as 'Ekman pumping' \citep{ekman:AMA1905},
and it is thus a key factor in the life-cycle of large-scale synoptic systems.
%
% %% This shear is probably not related to Ekman-turning 
% On the mesoscale, shear in the surface layer a crucial factor for the initiation and sustainment of strong convective events (REFERENCE?).
% 
Locally, the directional shear of the wind in the upper part of the surface layer may cause a systmatic yaw
for tall wind power generation devices where blades reach into the Ekman layer, i.e. that part of the boundary
layer where the wind starts to turn. 
%
\par
%
Wind veer in the planetary boundary layer is commonly characterized by the surface veering angle $\alpha$
defined as the angle between the negative surface shear stress $\tau_\SFC$ and the geostrophic wind.
%
Surface veering $\alpha$ and geostrophic drag $Z\equiv u_\star/G$, where $u_\star\equiv\sqrt{\nu|\tau_\SFC|}$,
uniquely determine the surface drag $\tau_\SFC$ in a turbulent Ekman flow. 
%
In any quantitative description of the surface layer, the friction velocity $u_\star$ is the dynamic scale
and $\alpha$ defines the alignment of the frame of reference.
%
Knowledge about $u_\star$ and $\alpha$ is thus a pre-requisite for any quantitative theory of the surface layer and a first
attempt in theoretical quantification was undertaken by \citet{rossby:PPO1935} using integral relations in the boundary layer.
%
Asymptotic similarity theory was later used by \citet{tennekes:JAS1973, blackadar:JAS1968},
and--based on his seminal direct numerical simulations (DNS) of Ekman flow--,~\cite{spalart:JFM1989} suggested
a modification to take into account effects of low to intermediate Reynolds numbers.
Later on, constants were re-evaluated with a focus on the atmospheric boundary layer based on
observations \cite{hogstrom:BM1988,hogstrom:BM1996}
and numerical modelling \citep{spalart:PF2008, spalart:PF2009, ansorge:BM2014, ansorge:BM2019}. 
%
\par
%
Attempts were also undertaken to obtain profiles of the wind speed: 
One approach is to match the inner and outer layer at a reference height;
\citet{etling:2002, emeis:2018}  (Sec. 21.10; Eq. 21.48) choose the Prandtl-layer height to match
the wind speed profiles, which, however, requires to externally prescribe the veering $\alpha$ at that height. 
A one-dimensional profile with constant veering is given by \citet[Sec. 3; Eq. 3.1-3.19]{emeis:m2007}.

\citet{gryning:BM2007} present an extension of the wind-speed profile beyond the surface layer

@@

\citet{jiang:JAS2018} recognized that the outer part of the Ekman boundary layer receives less attention in comparison

@@

Despite advances in analysis of this simplified set-up \citep{jiang:JAS2018} \todo{(MORE?)},
there is yet insufficient understanding for a quantitative generalization of the results to
arbitrary external forcing (manifest in variation of the Reynolds number).

@@

\todo{Numerical approach, DNS, Scale-separation, Re-similarity
  \citep{dimotakis:JFM2000, dasilva:ARF2014, pope:2000, moin:ARF1998, mellado:JAM2018}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Formulation and Numerical Approach}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We consider here incompressible, turbulent Ekman flow, that is, the turbulent flow over a flat rotating plate, as
a physical model for the truly neutral atmospheric boundary layer (ABL).
%
The f-plane approximation is applied such that rotation only acts on horizontal velocity components;
we thus neglect rotational effects on the horizontal components of velocity and dynamical effects
due to latitudinal variation of the rate of rotation.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation and Governing Equations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The dimensional velocity vector of the numerical simulations is $\underline{U} = (U_1,U_2,U_3) = (U,V,W)$ over the coordinate system
$Oxyz$, where an approximate alignment (plus/ minus few degrees centigrade) of the direction $Ox$ with the surface shear stress is achieved.
The coordinate $Oy$ points away from the wall,  and $Oz$ points in the span-wise direction normal to $Oxy$.
For analysis of the results, we use two coordinate systems that are
(i)  exactly aligned with the surface shear stress and labelled by an upper index $\alpha$ as in $\underline{U}^\alpha$ for the velocity vector, and
(ii) the coordinate system aligned with the free-atmosphere geostrophic wind labelled by an upper index $G$ as in $\underline{U}^G$.
%
We denote by $u_\star$ the modulus surface friction velocity and let $Z_\star=1/u_\star$ for brevity. 
%
\par
%
We consider the incompressible Navier--Stokes equations for the three velocity components on the f-plane in a framework
that is governed by the physical parameters
(i)   geostrophic wind magnitude $G=\sqrt{G_1^2+G_3^2}$ (where $\mathbf{G}=(G_1,0,G_3)$ is the geostrophic wind vector),
(ii)  Coriolis parameter $f$ (representing the angular rotation), and
(iii) kinematic viscosity $\nu$.
   %
These yield a Reynolds number $\RE$ as non-dimensional parameter governing the system.
%
We use the geostrophic wind as velocity and the Coriolis parameter $f$ as time scale for the non-dimensional framework which implies the Rossby radius $\LR=G/f$ as length scale, such that
\begin{equation}
  \RE=\frac{G \LR }{\nu}. 
\end{equation}
%
\par
%
The non-dimensionalized set of governing equation reads as 
\begin{subequations} 
\begin{eqnarray}
  \frac{\partial u_i}{\partial t} &=& \frac{\p \pi}{\p x_i} - u_j \frac{\p u_i}{\p x_j} + \epsilon_{i2j} (u_j -G_j)   + \frac{1}{\RE} \frac{\p^2 u_i}{\p x_{j}^2} \\ 
  \frac{\partial u_j}{\partial x_j} &=& 0,   
\end{eqnarray} 
\label{eqn:governing} 
\end{subequations}
where $u_i=U_i/G$ are the non-dimensional components velocity, $\pi$ is non-dimensional pressure, and $g_j=G_j/G$ are non-dimensionalized components geostrophic wind (with $g_{1}^2+g_{3}^2=1$ by construction). 
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Numerical Simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{DNS Code}
\par
\todo{(new and old) Simulations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Formulations for the outer layer that take into account the rotation, and thus deviation from the channel-flow analogy
need to be matched to the framework of surface similarity, and it
occurs that a smooth transition from the inner to the outer layer is not easily achieved.
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we believe that it is the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.
